\documentclass[12pt,a4paper]{article}
\usepackage[left=2cm,right=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{multirow}

\usepackage[english]{babel}
\usepackage{cite}

\usepackage{makeidx}
\makeindex

\usepackage{graphicx}
\graphicspath{{Pictures/}} 

\usepackage{amsmath}
\usepackage{amssymb}

\input{thm_header}


\title{High-Assurance Quasi-Synchronous Systems}
\date{}
\author{Robin Larrieu -- \'Ecole Polytechnique \\ Natarajan Shankar -- SRI International}

\begin{document}

%\newgeometry{left=3cm, right=3cm, top=4cm, bottom=2.5cm} % different margins for abstact + TOC
\maketitle

\begin{abstract}
The design of a complex cyber-physical system is centered around one
or more models of computation (MoCs).  These models define the
semantic framework within which a network of sensors, controllers, and
actuators operate and interact with each other.  In this paper, we
examine the foundations of a quasi-synchronous model of computation
Our version of the quasi-synchronous model is inspired by the Robot
Operating System (ROS).  It consists of nodes that encapsulate
computation and topic channels that are used for communicating between
nodes.  The nodes execute with a fixed period with possible jitter due
to local clock drift and scheduling uncertainties, but are not
otherwise synchronized.  The channels are implemented through a
mailbox semantics.  In each execution step, a node reads its incoming
mailboxes, applies a next-step operation to update its local state,
and writes to all its outgoing mailboxes.  The underlying transport
mechanism implements the mailbox-to-mailbox data transfer with some
bounded latency.  Messages can be lost if a mailbox is over-written
before it is read.  We prove a number of basic theorems that are
useful for designing robust high-assurance cyber-physical systems
using this simple model of computation.  We show that depending on the
relative rates of the sender and receiver, there is a bound on the
number of consecutive messages that can be lost.  By increasing the
mailbox queue size to a given bound, message loss can be eliminated.
We demonstrate that there is a bounded latency between the time a
message is sent and the time that it is processed, which in turn can
be used to bound the end-to-end sense-control-actuate latency.  We
illustrate how these theorems are useful in designing and verifying a
correct thermostat-based heating system.  Our proofs have been
mechanically verified using the Prototype Verification System (PVS). 
\end{abstract}
\newpage
 
%\renewcommand{\contentsname}{Table of Contents \vspace{1cm}}
\tableofcontents

\restoregeometry % return to default margins
\newpage

\section{Introduction}
\input{introduction}
\input{intro_contents_paper}
\newpage

\section{The ROS architecture}
\input{ROS_arch}

\section{Low-level messaging system}
\input{messages}


\section{Assurance claim for the plant controller}
\input{plant_controller}

%\section{Further work: obstacle avoidance}
%\input{obstacle_avoidance}


\bibliographystyle{plain}
\bibliography{biblio}


\end{document}
